Bayesian exponential random graph modeling of KWL Burial network

### 2020-07-17 analysis/paper/paper.Rmd
update code and edit the method section

```{r tidy-data}
# combine some periods
burial_three_period_age_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(ritual = case_when(
    `Stamped_ceramic` == "2" ~ "two pots",
    `Stamped_ceramic` == "1" ~ "one pot",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% #based on the result of histogram
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         ritual, # consider to remove if not very informative
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         Porcelain, #prestige good
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         quantity)

# number of each phase
burial_three_period_number <-
  burial_three_period_tidy %>%
  count(Phase)

pre_E <- burial_three_period_number$Phase[4]
post_E <- burial_three_period_number$Phase[3]

@@

```{r prep-network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

# create node list using burial index by phase
#--------------------pre--------------------------------
nodes_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  select(burial_label) %>%
  rowid_to_column("id")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_tidy %>%
  select(burial_label, 5:13) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_pre <-
map2(burial_comb_pre$burial_1,
     burial_comb_pre$burial_2,
     ~bind_cols(
       edge_list_pre %>%
       filter(burial_label == .x) %>%
       unnest(data),
       edge_list_pre %>%
         filter(burial_label == .y) %>%
         unnest(data)) %>%
       rowwise() %>%
       mutate(common_counts = sum(count, count1)))

@@

```{r prep-network-object-post}
# filter post burials
burial_post <-
  burial_three_period_tidy %>%
  filter(Phase == "post") %>%
  janitor::remove_empty(which = "cols")

# create node list
nodes_post <-
  burial_three_period_tidy %>%
  filter(Phase == "post") %>%
  select(burial_label) %>%
  rowid_to_column("id")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_post <-
  burial_three_period_tidy %>%
  select(burial_label, 5:13) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_post <-
  map2(burial_comb_post$burial_1,
       burial_comb_post$burial_2,
       ~bind_cols(
         edge_list_post %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_post %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count, count1)))

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014]. Bergm provides a framework for Bayesian parameter estimation on the result of exponential random graph models, which depends on ergm packages for simulation and model specification [@Hunter2008; @Morris2008; @Handcock2008]. Despite exponential random graph models can better describe a network by capturing the dependency structure, it is difficult to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a step to make probability distribution to integrate to one, which becomes hard to compute for larger set of networks [@Caimo2017bayesian]. Another problem is the degeneracy that means the probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. To overcome these limitations, Bayesian model applies Markov chain Monte Carlo (MCMC) simulation with the exchange algorithm, which solves computational problems in ERGMs [@Caimo2011].

@@

Markov chain Monte Carlo (MCMC) is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we seek, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within that chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical approach to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network. Our observed networks, pre-European and post-European networks, followed the normal distribution that is a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly observed in real world [@Caimo2017]. Since our original data implies higher transitivity due to potentially multiple groups having the same kind of burial good in common, we specified the prior based on our observation with high transitivity and low density.

@@

The pre-European network shows three subgroups that possessing the same types of burial goods: medium amount of carnelian bead on the right, medium amount of gold-foil beads on the left, and porcelains in the middle. Although burials with different amount of burial goods can be identified, there is no clear clusters demonstrating the sharing of prestige goods is limited to burials with rich goods.

>>>>

```{r tidy-data}
# combine some periods
burial_three_period_age_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(ritual = case_when(
    `Stamped_ceramic` == "2" ~ "two pots",
    `Stamped_ceramic` == "1" ~ "one pot",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% #based on the result of histogram
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         ritual, # consider to remove if not very informative
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         #Agate_bead, #female burials
         #Golden_bead,
         Porcelain, #prestige good
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         #Bell, #children's burials
         quantity)

# number of each phase
burial_three_period_age_number <-
  burial_three_period_age_tidy %>%
  count(Phase)

pre_E <- burial_three_period_age_number$Phase[4]
post_E <- burial_three_period_age_number$Phase[3]

@@

```{r prep-network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_age_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

# create node list using burial index by phase
#--------------------pre--------------------------------
nodes_pre <-
  burial_three_period_age_tidy %>%
  filter(Phase == "pre") %>%
  select(burial_label) %>%
  rowid_to_column("id")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_age_tidy %>%
  select(burial_label, 6:14) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_pre <-
  map2(burial_comb_pre$burial_1,
       burial_comb_pre$burial_2,
       ~bind_cols(
         edge_list_pre %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_pre %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count...3, count...6)))

@@

```{r prep-network-object-post}
# filter post burials
burial_post <-
  burial_three_period_age_tidy %>%
  filter(Phase == "post") %>%
  janitor::remove_empty(which = "cols") %>%
  filter(quantity != "none") # remove burial without burial goods

# create node list
nodes_post <-
  burial_three_period_age_tidy %>%
  filter(Phase == "post") %>%
  filter(quantity != "none") %>%
  select(burial_label) %>%
  rowid_to_column("id")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_post <-
  burial_three_period_age_tidy %>%
  select(burial_label, 6:14) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_post <-
  map2(burial_comb_post$burial_1,
       burial_comb_post$burial_2,
       ~bind_cols(
         edge_list_post %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_post %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count...3, count...6)))

@@

We analyzed social networks using the R programming language with bergm package for Bayesian exponential random graph models [@Caimo2014]. Bergm provides a framework for Bayesian parameter estimation on the result of exponential random graph models, which depends on ergm packages for simulation and model specification [@Hunter2008; @Morris2008; @Handcock2008]. Although exponential random graph models can better describe a network by capturing the dependency structure, it is difficult to estimate model parameters and interpret the result due to the intractable normalizing constant and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a step to make probability distribution to integrate to one, which becomes hard to compute for larger set of networks [@Caimo2017bayesian]. Another problem is the degeneracy that means the probability models tend to overestimate a small number of extreme graphs by assigning too much weight, such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. To overcome these limitations, Bayesian model applies Markov chain Monte Carlo (MCMC) simulation with the exchange algorithm, which solves computational problems in ERGMs [@Caimo2011].

@@

Markov chain Monte Carlo (MCMC) is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we study, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical approach to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network. Our observed networks, pre-European and post-European networks, followed the normal distribution that is a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly observed in real world [@Caimo2017]. Since our original data implies higher transitivity due to potentially multiple groups having the same kind of burial good in common, we specified the prior based on our observation with high transitivity and low density.

@@

The pre-European network shows three subgroups that possessing the same types of burial goods: medium amount of carnelian bead on the right, medium amount of gold-foil beads on the left, and porcelains in the middle. Although burials with different amount of burial goods can be identified, there is no clear clusters demonstrating the sharing of prestige goods only for burials with rich goods.
